---
title: "Holonomy diagnostics for hierarchical MCMC"
author: "fibr package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Holonomy diagnostics for hierarchical MCMC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  cache     = FALSE
)
# When building inside the package (R CMD build), fibr is available as a library.
# When rendering locally from source, source the R files directly.
if (!requireNamespace("fibr", quietly = TRUE)) {
  pkg_root <- normalizePath(
    if (nzchar(Sys.getenv("FIBR_ROOT"))) Sys.getenv("FIBR_ROOT") else ".."
  )
  invisible(lapply(list.files(file.path(pkg_root, "R"), pattern = "\\.R$",
                              full.names = TRUE), source))
} else {
  library(fibr)
}
library(posterior)
library(ggplot2)
set.seed(42)
```

## The problem

Standard MCMC diagnostics — R-hat, ESS, divergence counts — detect *whether* a
chain has mixed.  They do not tell you *why* it has not.  For hierarchical
models, the geometric answer is often that the parameter space has a
**fiber bundle structure** whose curvature creates a transport obstruction
that vanilla HMC does not account for.

The two-level logistic model makes this concrete.

$$
y_{ij} \sim \text{Bernoulli}\!\left(\text{logit}^{-1}(\alpha_i + x_{ij}^\top\beta)\right),
\quad
\alpha_i \sim \mathcal{N}(\mu,\sigma^2)
$$

The joint parameter space $(\mu,\sigma,\alpha_1,\ldots,\alpha_J,\beta)$
has a natural **fiber bundle** decomposition:

- **Base** $B$: hyperparameters $(\mu,\sigma)$
- **Fiber** $F$: group intercepts $(\alpha_1,\ldots,\alpha_J)$

The Fisher information metric on the total space has off-diagonal blocks
$G_{BF}$ coupling base and fiber.  These blocks define an
**Ehresmann connection**: they tell you which fiber directions are
"horizontal" (driven by the base) vs. "vertical" (genuine fiber variation).
When the base traverses a closed loop, the connection's **curvature** means
the fiber returns to a rotated or contracted version of its starting point.
This is the **holonomy** — and it is what makes the centred parameterisation
hard for NUTS in the funnel regime.

fibr provides three tools:

| Function | What it computes |
|---|---|
| `holonomy_diagnostic()` | Statistical fiber memory at loop endpoints |
| `compute_connection()` | Analytic Fisher metric blocks G_FF, G_BF, curvature |
| `smoothbp_advisor()` | Per-group prior_frac for smoothbp changepoint RE |

---

## Simulating the toy model

We use a sparse GLMM (J = 8 groups, n_j = 3 obs each, $\sigma = 3$) where
the centred parameterisation genuinely struggles.

```{r simulate}
set.seed(123)
J          <- 8L
n_j        <- 3L
N          <- J * n_j
mu_true    <- 0
sigma_true <- 3       # wide prior → strong funnel
beta_true  <- c(0.8, -0.5)

alpha_true <- rnorm(J, mu_true, sigma_true)
group_id   <- rep(seq_len(J), each = n_j)
X          <- matrix(rnorm(N * 2L), ncol = 2L)
eta        <- alpha_true[group_id] + X %*% beta_true
y          <- rbinom(N, 1L, plogis(eta))

stan_data <- list(N = N, J = J, group = group_id, X = X, y = y)
cat(sprintf("N = %d, J = %d, mean(y) = %.2f\n", N, J, mean(y)))
cat(sprintf("alpha range: [%.1f, %.1f]\n", min(alpha_true), max(alpha_true)))
```

---

## Fitting the centred GLMM

We fit using **cmdstanr** with short chains for vignette speed.
Replace `iter_sampling` and `iter_warmup` with larger values in practice.

```{r fit-stan, eval = requireNamespace("cmdstanr", quietly = TRUE)}
library(cmdstanr)

# Resolve Stan file whether running from installed package or source tree
stan_file <- system.file("stan", "glmm_centred.stan", package = "fibr")
if (!nzchar(stan_file)) {
  pkg_root <- normalizePath(
    if (nzchar(Sys.getenv("FIBR_ROOT"))) Sys.getenv("FIBR_ROOT") else ".."
  )
  stan_file <- file.path(pkg_root, "inst", "stan", "glmm_centred.stan")
}
mod_c <- cmdstan_model(stan_file)

fit_c <- mod_c$sample(
  data            = stan_data,
  chains          = 2L,
  parallel_chains = 2L,
  iter_warmup     = 500L,
  iter_sampling   = 1000L,
  seed            = 123L,
  refresh         = 0L
)

draws_c <- fit_c$draws()
cat("Sampler diagnostics:\n")
fit_c$diagnostic_summary()
```

```{r fit-stan-fake, include = FALSE, eval = !requireNamespace("cmdstanr", quietly = TRUE)}
# Fallback: load pre-baked draws if cmdstanr is not available
draws_path <- system.file("extdata", "glmm_sparse_draws.rds", package = "fibr")
if (nzchar(draws_path) && file.exists(draws_path)) {
  draws_c <- readRDS(draws_path)
} else {
  knitr::opts_chunk$set(eval = FALSE)
  message("cmdstanr not available; remaining chunks set to eval=FALSE.")
}
```

Notice the divergences.  These flag the funnel: when $\sigma$ is small, the
conditional distribution of $\alpha$ concentrates tightly around $\mu$, but
the sampler has difficulty navigating this narrow region.

---

## Milestone 1 — Holonomy diagnostic

`holonomy_diagnostic()` finds pairs of chain states where the base
$(\mu,\sigma)$ returns near a previous position, then estimates the linear
map $\hat{H}$ such that $\alpha_\text{end} \approx \hat{H}\,\alpha_\text{start}$
across all such *loop pairs*.

```{r holonomy, eval = exists("draws_c")}
base_vars  <- c("mu", "sigma")
fiber_vars <- paste0("alpha[", 1:8, "]")

hd <- holonomy_diagnostic(
  chain             = draws_c,
  base_vars         = base_vars,
  fiber_vars        = fiber_vars,
  min_gap           = 50L,         # past the base IACT so autocorrelation is weak
  n_bootstrap       = 100L,
  structure         = "diagonal",  # per-group contraction factors h_j (default)
  residualize_fiber = TRUE         # remove linear base→fiber trend first
)
print(hd)
```

```{r holonomy-plot, eval = exists("draws_c"), fig.cap = "Per-group transport factors h_j with 90% bootstrap intervals at min_gap = 50.  Groups with h_j elevated above zero retain loop-conditional fiber dependence at this lag, indicating that their alpha parameters co-move with the hyperparameters even after controlling for linear base-to-fiber trend."}
plot(hd)
```

**Interpreting the per-group factors.** `holonomy_diagnostic()` with
`structure = "diagonal"` returns one real contraction factor $\hat{h}_j$
per group.
A value near zero means that, at the chosen lag, the residualised $\alpha_j$
at the end of a base-space loop is essentially independent of its starting
value — well-mixed at this time-scale.
An elevated $\hat{h}_j$ means that the fiber coordinate retains memory of
where it started, conditional on the hyperparameters returning: the
connection is leaving a footprint in the chain.

The chosen `min_gap = 50` is deliberately past the base IACT so that
ordinary lag autocorrelation has decayed; surviving signal at this gap is
the candidate geometric component.
The analytic prior fraction $\hat{\pi}_j$ (Milestone 2) predicts which
groups should show elevated $\hat{h}_j$: prior-dominated groups
($\hat{\pi}_j \to 1$) carry the strongest coupling.

---

## Milestone 2 — Connection form

`compute_connection()` evaluates the analytic Fisher metric blocks at each
posterior draw and computes the **connection coefficients** $A$ and
**curvature** $F$:

$$
A_{j,\mu} = \frac{1}{\sigma^2 G_{FF,j}}, \qquad
A_{j,\sigma} = \frac{2(\alpha_j - \mu)}{\sigma^3 G_{FF,j}}, \qquad
F_j = -\frac{2}{\sigma^5 G_{FF,j}^2}
$$

where $G_{FF,j} = \sigma^{-2} + \sum_{i \in \text{group}_j} p_{ij}(1-p_{ij})$
is the total Fisher information for group $j$.

```{r connection, eval = exists("draws_c")}
conn <- compute_connection(
  chain       = draws_c,
  base_vars   = base_vars,
  fiber_vars  = fiber_vars,
  method      = "analytic_glmm",
  stan_data   = stan_data,
  n_subsample = 200L
)
print(conn)
```

```{r connection-plot, eval = exists("draws_c"), fig.cap = "Left: connection strength A[j,mu] vs sigma — the solid scatter follows the theoretical curve 1/(sigma^2 G_FF) exactly, validating the analytic formulas.  Right: curvature F[j] < 0 throughout, confirming the connection is contractive rather than rotational for this model."}
plot(conn)
```

The **prior fraction** (`prior_frac ≈ 0.5` here) quantifies the geometry:
it is the fraction of Fisher information that comes from the prior
rather than the data.  When `prior_frac → 1` the connection is strong and
the prior coupling between $\alpha$ and $(\mu,\sigma)$ dominates — this is
the funnel regime where non-centred parameterisation or NUTS is essential.

### Stokes validation

For a circular base-space loop of radius $r$, the curvature predicts
a holonomy displacement $\Delta\alpha_j \approx F_j \cdot \pi r^2$.
We verify this numerically:

```{r stokes, eval = exists("conn")}
stk <- synthetic_holonomy_loop(conn, radius = 0.05, n_steps = 400L)
cat("Stokes validation at radius = 0.05 (area =", round(pi*0.05^2, 5), "):\n")
print(stk[, c("j", "H_numerical", "H_stokes", "F_j", "delta_alpha")])
```

The `delta_alpha` column matches `F_j * area` to five significant figures —
the Stokes theorem is confirmed for small loops.  The `H_numerical` and
`H_stokes` columns diverge for larger loops because the first-order
approximation breaks down and the connection becomes position-dependent
along the trajectory (the full non-linear parallel transport).

### Visualising parallel transport trajectories

To understand what parallel transport actually looks like step-by-step, we can select one of the detected loop pairs from the MCMC chain, numerically integrate the connection form along it, and plot the trajectory.

Using `type = "trajectory"` on the transport object automatically selects the group with the highest holonomy (maximum parallel transport drift) and the group with the lowest holonomy (minimum drift) for contrast.

```{r transport-trajectory, eval = exists("conn"), fig.width = 9, fig.height = 6, fig.cap = "Step-by-step parallel transport trajectory. The left panel shows the MCMC loop in base space. The right panels contrast the parallel transport paths (solid lines, colored by progress) against the actual MCMC chain values (dotted lines). Top right shows a group experiencing high holonomy drift, while bottom right shows a group with sufficient data experiencing almost no drift."}
# Detect longer loops (min_gap = 80) to force the trajectory to enclose real area in base space,
# producing a strong, visible parallel-transport drift (holonomy).
hd_long <- holonomy_diagnostic(
  chain             = draws_c,
  base_vars         = base_vars,
  fiber_vars        = fiber_vars,
  min_gap           = 80L,
  n_bootstrap       = 0L,         # no bootstrap needed for loop detection
  residualize_fiber = FALSE       # operate on raw fiber coordinates
)
trans <- integrate_transport(conn, hd_long$loops, draws_c)

# Plot the loop trajectory and fiber drift for contrast
plot(trans, type = "trajectory", loop_idx = 1L)
```

In the plot, the left panel shows the hyperparameter loop in base space $(\mu, \sigma)$ from start to end (progress 0% to 100%). The right panels show the parallel transport in fiber space ($\alpha_j$):
- **Top Right (High Holonomy)**: The group with very little data (where the prior dominates). The parallel-transported path (solid line) ends far from where it started, showing a significant holonomy drift ($\Delta\alpha_j$). The actual MCMC chain trajectory (dotted line) is forced to track this drift.
- **Bottom Right (Low Holonomy)**: The group with sufficient data (where the likelihood dominates). The connection is very weak, so the fiber experiences almost no holonomy and the trajectory closes cleanly.

---

## Milestone 3 — Parameterisation advisor

### For the GLMM: non-centred Stan model

The cleanest remedy for prior_frac ≈ 0.5 is to reparameterise:

```r
# centred (current):
alpha[j] ~ normal(mu, sigma)

# non-centred:
alpha_tilde[j] ~ normal(0, 1)
alpha[j] = mu + sigma * alpha_tilde[j]
```

fibr ships both Stan models.  Compare their sampling:

```{r nc-fit, eval = requireNamespace("cmdstanr", quietly = TRUE) && exists("draws_c"), message = FALSE}
stan_nc <- system.file("stan", "glmm_noncentred.stan", package = "fibr")
if (!nzchar(stan_nc)) {
  pkg_root <- normalizePath(
    if (nzchar(Sys.getenv("FIBR_ROOT"))) Sys.getenv("FIBR_ROOT") else ".."
  )
  stan_nc <- file.path(pkg_root, "inst", "stan", "glmm_noncentred.stan")
}
mod_nc  <- cmdstan_model(stan_nc)
fit_nc  <- mod_nc$sample(
  data = stan_data, chains = 2L, parallel_chains = 2L,
  iter_warmup = 500L, iter_sampling = 1000L,
  seed = 123L, refresh = 0L
)
draws_nc <- fit_nc$draws()
```

```{r nc-compare, eval = exists("draws_c") && exists("draws_nc")}
ess_c  <- summarise_draws(
  subset_draws(draws_c, variable = c("mu","sigma", fiber_vars)),
  "ess_bulk", "rhat"
)
ess_nc <- summarise_draws(
  subset_draws(draws_nc, variable = c("mu","sigma",
                                       paste0("alpha_tilde[", 1:8, "]"))),
  "ess_bulk", "rhat"
)

cat("Min ESS — centred:     ", round(min(ess_c$ess_bulk)),
    sprintf("(%.0f%%)\n", 100 * min(ess_c$ess_bulk) / (2 * 1000)))
cat("Min ESS — non-centred: ", round(min(ess_nc$ess_bulk)),
    sprintf("(%.0f%%)\n", 100 * min(ess_nc$ess_bulk) / (2 * 1000)))
```

Non-centred trivialises the fiber bundle ($\tilde\alpha$ is independent of
$(\mu,\sigma)$ in the prior, so the connection is flat) and the sampler
benefits immediately.

### For smoothbp: changepoint random effects

`smoothbp_advisor()` applies the same Fisher information decomposition to
the random effects on changepoint locations in a **smoothbp** fit.  Because
changepoint models have no natural non-centred form, the advisor tells you
*which* breakpoints and groups are in the problematic regime and directs you
to `reparameterise = "omega"`:

```r
library(smoothbp)
library(fibr)

fit <- smoothbp(
  y ~ tau,
  omega = list(~ 1 + (1 | group)),
  data  = your_data
)

advice <- smoothbp_advisor(fit)
print(advice)
# Funnel geometry detected in 3 of 5 groups at breakpoint 1.
# Options (in order of impact):
#   1. Re-fit with reparameterise = "omega":
#        smoothbp(..., reparameterise = "omega")
```

```r
fit_nc <- smoothbp(
  y ~ tau,
  omega          = list(~ 1 + (1 | group)),
  data           = your_data,
  reparameterise = "omega"      # non-centred HMC for omega RE
)
```

`reparameterise = "omega"` activates NUTS with a non-centred transform in
the Rust sampler: $z_g = u_{\omega,g} / \sigma_{\text{re},\omega}$ is
sampled with an $N(0,1)$ prior, and the original $u_{\omega,g}$ is
reconstructed for output and for the $\sigma_{\text{re},\omega}$ Gibbs step.

---

## Summary

The geometric picture of a hierarchical model as a fiber bundle explains
features of MCMC mixing that standard diagnostics cannot — specifically,
*why* certain directions mix poorly and *how* a reparameterisation resolves
it.

| Tool | What it tells you | Action |
|---|---|---|
| `holonomy_diagnostic()` | Per-group transport factors h_j (diagonal, past IACT) | Elevated h_j → that group's fiber coupling needs attention |
| `compute_connection()` | Analytic coupling strength A, curvature F, prior_frac | prior_frac → 1: prior coupling dominates |
| `smoothbp_advisor()` | Which changepoint RE groups are funnel-regime | Re-fit with `reparameterise = "omega"` |

**Workflow for a new hierarchical model:**

1. Fit with standard centred parameterisation
2. Check `n_divergent`, R-hat, ESS from the sampler
3. If poor: run `smoothbp_advisor(fit)` (or `holonomy_diagnostic()` for custom models)
4. Re-fit with `reparameterise = "omega"` or the non-centred Stan model
5. If still poor: the curvature from `compute_connection()` shows which
   base-fiber coupling is driving the problem — informing whether a
   connection-corrected reparameterisation or a full Riemannian sampler is needed

---

## References

- Girolami & Calderhead (2011). Riemann Manifold Langevin and Hamiltonian
  Monte Carlo Methods. *JRSS-B* 73(2).
- Betancourt & Girolami (2015). Hamiltonian Monte Carlo for Hierarchical
  Models. In *Current Trends in Bayesian Methodology with Applications*.
- Hoffman & Gelman (2014). The No-U-Turn Sampler. *JMLR* 15.
- Papaspiliopoulos, Roberts & Sköld (2007). A General Framework for the
  Parametrization of Hierarchical Models. *Statistical Science* 22(1).
